---
title: "Medication Adherence: Training Material"
author: "Martin Ng"
output: 
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r seed}
set.seed(41)
```

```{r libraries}
library(data.table)
library(DT)
```

```{r constants}

```

```{r functions}
round.numerics <- function(x, digits){
  if(is.numeric(x)){
    x <- round(x = x, digits = digits)
  }
  return(x)
}
```

```{r load_data}
adherence <- fread(file.path("..", "data", "adherence.csv"), verbose = F)  
baseline_measurements <- fread(file.path("..", "data", "baseline_measurements.csv"), verbose = F)
```

```{r explore_data, eval = FALSE}

```


```{r clean_data}

```



## Introduction {.tabset}

This is the training material for generating a monthly report for health insurance company X, intended to highlight some of the most challenging parts of understanding and utilizing the data.

The components to generate this report include:

* A reporting template (.Rmd file) that is programmed to calculate the answers to the questions provided, this serves as the basic monthly report.
* A reporting template (.Rmd file) that is programmed to perform some independent investigation.


### Data Exploration {.tabset}

Files needed for generating report include:

* Baseline Measurements e.g. “baseline measurements.csv”
* Adherence e.g. “adherence.csv”


#### Baseline Measurements

Baseline Measurements contains baseline demographic and clinical information about patients.

All columns of data are correctly formatted with no missing value, so no cleaning needed.


```{r}
head(baseline_measurements)
```


#### Adherence 

Adherence contains information on patients' medication adherence over time.

All columns of data are correctly formatted with no missing value, so no cleaning needed.

```{r}
head(adherence)
```

Important concepts:

* t1: The starting time (in days) for a particular period of observation or adherence to a medication.
* e.g. For the first record, t1 is 0, meaning this is the first day of observation.

* t2: The ending time (in days) for the period of observation or adherence to a medication.
* e.g. For the first record, t2 is 32, meaning the observation period lasted for 32 days.

* ace, bb and statin: A binary variable indicating whether the patient was taking the med during this time period (1 for yes, 0 for no).



### Basic Report {.tabset}

The reporting template contains 10 questions where answers will be automatically generated.

Functions are created and used repeatedly to answer the questions, they are stored in the “functions” section.

Areas to be aware of:

#### Functions

**calculate_adherence_days()**:

To calculate adherence days for adherence rate, get each patient's t1 when medication is received (=1), which is saved as first_t1, then filter t1 >= first_t1 and get the sum of (t2-t1)* get(medication).
```{r, eval=FALSE, echo=TRUE}
# Function to calculate adherence days for each medication starting from the first adherence t1 for each patient
calculate_adherence_days <- function(adherence_data, first_data, medication) {
  # Merge the first adherence time back with the main adherence data within the first year
  merged_data <- merge(adherence_data, first_data[, .(id, first_t1 = t1)], by = "id", all.x = TRUE)
  
  # Filter the rows where t1 is greater than or equal to the first time the medication was possessed
  adherence_filtered <- merged_data[t1 >= first_t1]
  
  # Calculate the total adherence days for each patient
  adherence_days_summary <- adherence_filtered[, .(adherence_days = sum((t2 - t1) * get(medication))), by = id]
  
  return(adherence_days_summary)
}
```


**calculate_continuous_duration()**:

To calculate how long do patients continuously have a filled prescription, one must consider the case that no gap ever occurs.
```{r, eval=FALSE, echo=TRUE}
# If no gap occurs, set the end_t1 to the max of t2 (follow-up end)
if (length(end_t1) == 0) {
  end_t1 <- max(t2)
} else {
  end_t1 <- min(end_t1, na.rm = TRUE)
}
```



#### "One Year"

Both **Q2** and **Q4** ask for one-year adherence while **Q3** limits the answer to during the first year, so only use data where t2 <= 365, achieved by filtering t1 < 365 and then using pmin() to clip t2 at 365.
```{r, eval=FALSE, echo=TRUE}
# Keep only records within the first year (t2 <= 365)
adherence_1_year <- adherence_filtered[t1 < one_year]
adherence_1_year[, t2 := pmin(t2, one_year)] 
```

**Special Case**:

**Q9** asks for one-year adherence too, but **starting at two weeks after the initial diagnosis**, so only use data where t1>= 14 and t2<= 379 (365+14), achieved by filtering t1 < 379, then using pmax() to clip t1 at 14 and pmin() to clip t2 at 379.
```{r, eval=FALSE, echo=TRUE}
# Clip t1 to 14 and t2 to 379, but exclude records where t2 < 14
adherence_1_year <- adherence_filtered[t2 >= two_weeks & t1 < 379]    
adherence_1_year[, t1 := pmax(t1, two_weeks)]              
adherence_1_year[, t2 := pmin(t2, 379)]            
```



#### Q6

To conduct chi-squared test of independence, convert the data into a contingency table first.
```{r, eval=FALSE, echo=TRUE}
# Create a contingency table for the baseline condition variable
contingency_table <- table(merged_data$filled_statin, merged_data$baseline.condition)

# Perform chi-squared test
chi_squared_result <- chisq.test(contingency_table)
```


#### Q7

After conducting the logistic regression, the model coefficients are not the odds ratios. The estimated odds ratios are the exponential of the model's estimated coefficients. 
```{r, eval=FALSE, echo=TRUE}
# Extract odds ratios for each model
odds_ratios_ace <- exp(coef(model_ace))
odds_ratios_bb <- exp(coef(model_bb))
odds_ratios_statin <- exp(coef(model_statin))
```




### Independent Investigation {.tabset}

The independent investigation focused on 2 areas:

1. Differences in covariates between early initiators and non-initiators for **all medications**.

2. Effect of polypharmacy on the adherence of patients


#### Early vs Non-initiators

For the t-tests and Chi-squared tests for ACE Inhibitors and Beta Blockers, reference Q6 and change variable names.

#### Polypharmacy

Since one-year adherence is examined here, dont forget to clip the t2 at 365 like Q2-4.
```{r, eval=FALSE, echo=TRUE}
# Filter patients with at least 1 year of follow-up
adherence_filtered <- filter_follow_up(adherence, one_year)
adherence_1_year <- adherence_filtered[t1 < one_year]
adherence_1_year[, t2 := pmin(t2, one_year)] # Clip t2 at 365 days to stay within the first year
```










